Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-05-24 papers/paper-eccentricity/paper-eccentricity.tex

\subsection{Reproducing the eccentricity scatter with a toy model}\label{ssec:toy}

We can gain insight into the large binary eccentricity scatter observed in the
runs with high galactic merger orbit eccentricities by considering a simplified
model that captures the essential features of how the SMBH binary becomes bound
and settles on an orbit with a given eccentricity.
The components we include in this model are the two SMBHs,
a smooth background potential, and dynamical friction.
For simplicity, we consider only symmetrical 1:1 mergers, and take the BH orbits
to lie in the $xz$-plane.
In this case the relative motion of the BHs can be treated as a one-body problem,
with the equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}$ is the separation vector of the BHs, $M_\bullet$ the mass of a single BH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the background potential,
and $\vb{a}_\mathrm{DF}$ the acceleration due to dynamical friction.

The background galaxy is modelled as a with constant density spheroidal potential,
causing the acceleration \citep[e.g.][]{binney2008}
\begin{equation}
\vb{a}_\mathrm{bg} = - 2 \pi G \rho \operatorname{diag}(A_x,A_y,A_z) \vb{x},
\end{equation}
where $\rho$ is the stellar density,
and the $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{equation}
A_x = A_y  = 2 \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]
\end{equation}
\begin{equation}
A_z = \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right].
\end{equation}
With eccentricity values of $e_\mathrm{s} \approx 0.9$,
this spheroidal potential matches the central potential in the galaxy during the
BH binary binding phase reasonably well,
apart from the contributions from the BHs themselves and the groups of stars bound to them.

For dynamical friction, we use the \drafting{Chandrasekhar ref} formula assuming
a Maxwellian distribution with a constant velocity dispersion $\sigma$.
This formula gives the acceleration of a single BH due to dynamical friction as
\begin{equation}\label{eq:chandra_df}
\vb{a}_{1,\mathrm{DF}} = - 4 \pi G^2 M_\bullet \rho \ln{\Lambda} 
                        (\operatorname{erf}(X) - 2 \pi^{-1/2} X e^{-X^2}) \frac{\vb{v}_{1}}{|\vb{v}_{1}|^3},
\end{equation}
where $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH,
$X = |\vb{v}_1|/\sqrt{2}\sigma$,
and the Coulomb logarithm is taken to have the value $\ln{\Lambda} = 10$.
To account for differences between the constant density background of stars
assumed here and the more complex system that is present in the simulations
we additionally multiply the acceleration given by Eq. \eqref{eq:chandra_df}
with a scaling and cut-off function
\begin{equation}
f(a) = \frac{\alpha}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]},
\end{equation}
which serves to remove the effect of dynamical friction when the BH binary has become
hard in addition to allowing the strength of dynamical friction to be scaled to
achieve similar behaviour as observed in the simulations.
Here $a$ is the semi-major axis of a bound binary, and for unbound binaries we set $f=\alpha$.
We set the scaling factor $\alpha=0.5$, and use cut-off scales $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ being the semimajor-axis of a hard BH binary.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

After specifying the initial position and velocity, as well as the other parameters of the model,
the equation of motion can be solved using standard numerical integration techniques.
We use the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}
for this purpose.
To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xz$-plane as 
\begin{gather}
\vb{x}_0 = (r_0, b)\\
\dot{\vb{x}}_0 = (-v_0, 0).
\end{gather}
We fix $r_0 = 25\,\mathrm{pc}$,
and use model parameters
$\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
and
$M_\bullet=10^8\,\Msun$,
which are fairly similar to the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.90$ and $e_0=0.99$ merger orbit simulations.
Based on the simulations we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.90$ case,
and $v_0 = 560\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.99$ case.
The spheroid eccentricity parameter is not easily measurable from the simulations,
but by testing different values we find that $e_\mathrm{s}=0.905$ gives behaviour
that closely resembles the simulation results.
We compute the resulting eccentricities for a range of impact parameters up to $b=20\mathrm{pc}$
for the two initial velocities.
In the model, a $90\degr$ deflection during the first pericenter passage
corresponds to initial impact parameters of $b\approx 5.8\,\mathrm{pc}$ and
$b\approx 3.3\,\mathrm{pc}$ for the $e_0=0.90$ and $e_0=0.99$ cases, respectively.

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
In addition to the reference $e_\mathrm{s}=0.905$ model, the figure also shows the model results for
$e_\mathrm{s}=0.900$ and $e_\mathrm{s}=0.910$, as particularly the $e_0=0.99$ case is quite sensitive to this parameter.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
Otherwise the model appears to give a very good match to the simulation results,
and the required slight shift could be due to the galactic potential being slightly tilted in the simulations 
compared to the BH trajectories due to the complex dynamics at this stage.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta \gtrsim 120\degr$ region,
although the model does appear to capture the mean behaviour fairly well.
The behaviour of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.
Overall we can however see that this simple model correctly captures the behaviour of eccentricity seen in the simulations.

>>>>

\subsection{Reproducing the eccentricity behavior with a toy model}\label{ssec:toy}

To gain insight into the origin of the dependence of binary eccentricity on the 
deflection angle $\theta$ during the binary formation,
we have constructed a simple toy model that can qualitatively reproduce the observed
behavior from the simulations.
This model includes only the essential components relevant to this process,
which are the asymmetric background stellar potential,
dynamical friction from the stellar background,
and naturally the SMBHs themselves.

For simplicity, we consider only symmetrical 1:1 mergers, and take the BH orbits
to lie in the $xy$-plane.
In this case the relative motion of the BHs can be treated as a one-body problem,
with the equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the BHs, $M_\bullet$ the mass of a single BH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the background potential,
and $\vb{a}_\mathrm{DF}$ the acceleration due to dynamical friction.

The background galaxy is modeled as a constant density spheroidal potential,
causing the acceleration \citep[e.g.][]{binney2008}
\begin{equation}
\vb{a}_\mathrm{bg} = - 2 \pi G \rho \operatorname{diag}(A_x,A_y) \vb{x},
\end{equation}
where $\rho$ is the stellar density,
and the $A$ coefficients are related to the eccentricity $e_\mathrm{s}$ of the spheroid as
\begin{align}
A_x &= 2 \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
    \left[\frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right) - 1 \right]\\
A_y &= \frac{1-e_\mathrm{s}^2}{e_\mathrm{s}^2} 
        \left[ \frac{1}{1-e_\mathrm{s}^2} 
        - \frac{1}{2 e_\mathrm{s}} \ln\left(\frac{1+e_\mathrm{s}}{1-e_\mathrm{s}}\right)\right].
\end{align}
Here we have chosen the long axis of the spheroid to be aligned with the $x$-axis.
With eccentricity values of $e_\mathrm{s} \approx 0.9$,
this spheroidal potential matches the central potential in the galaxy during the
SMBH binary binding phase reasonably well,
apart from the contributions from the BHs themselves and the groups of stars bound to them.

For dynamical friction, we use the \citet{1943ApJ....97..255C} formula assuming
a Maxwellian distribution with a constant velocity dispersion $\sigma$ \citep[e.g.][]{binney2008}.
This formula gives the acceleration of a single BH due to dynamical friction as
\begin{equation}\label{eq:chandra_df}
\begin{aligned}
\vb{a}_{1,\mathrm{DF}} = &-\frac{4 \pi G^2 M_\bullet \rho \ln{\Lambda}}{|\vb{v}_{1}|^3}\\
                        &\times\left(\operatorname{erf}(X) - \frac{2X}{\sqrt{\pi}} \exp(-X^2)\right)\vb{v}_{1},
\end{aligned}
\end{equation}
where $\vb{v}_1 = \dot{\vb{x}}/2$ is the velocity of a single BH,
$X = |\vb{v}_1|/\sqrt{2}\sigma$,
and the Coulomb logarithm is taken to have the value $\ln{\Lambda} = 10$.
To account for differences between the constant density background of stars
assumed here and the more complex system that is present in the simulations
we additionally multiply the acceleration given by Eq.~\eqref{eq:chandra_df}
with a scaling and cut-off function
\begin{equation}
f(a) = \frac{\alpha}{1 + \exp[(a_\mathrm{c}-a)/d_\mathrm{c}]},
\end{equation}
which serves to remove the effect of dynamical friction when the SMBH binary has become
hard in addition to allowing the strength of dynamical friction to be scaled to
achieve similar behavior as observed in the simulations, since in the simulations the
stellar background is much more complex.
Here $a$ is the semi-major axis of a bound binary, and for unbound binaries we set $f=\alpha$.
We set the scaling factor $\alpha=0.5$, and use cut-off scales $a_\mathrm{c}=2 a_\mathrm{h}$
and $d_\mathrm{c} = 0.5 a_\mathrm{h}$,
with $a_\mathrm{h}$ being the semimajor-axis of a hard BH binary.
The total dynamical friction term is then
\begin{equation}
\vb{a}_{\mathrm{DF}} = 2 f(a) \vb{a}_{1,\mathrm{DF}}.
\end{equation}

To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xy$-plane as 
\begin{align}
\vb{x}_0 &= (r_0, b)\\
\dot{\vb{x}}_0 &= (-v_0, 0).
\end{align}
We fix $r_0 = 25\,\mathrm{pc}$,
and use model parameters
$\rho = 300 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
and
$M_\bullet=10^8\,\Msun$,
which are fairly similar to the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.90$ and $e_0=0.99$ merger orbit simulations.
Based on the simulations we set $v_0 = 450\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.90$ case,
and $v_0 = 560\,\mathrm{km\,s^{-1}}$ to match the $e_0=0.99$ case.
The spheroid eccentricity parameter is not easily measurable from the simulations,
and we therefore perform calculations with different values of $e_\mathrm{s}$.
We compute the resulting eccentricities for a range of impact parameters up to $b=20\,\mathrm{pc}$
for the two initial velocities,
by solving the equation of motion \eqref{eq:toy_model_eom} until the binary has become hard using
the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}.
The range of impact parameters covers most of the relevant parameter space,
as a $90\degr$ deflection during the first pericenter passage
corresponds to initial impact parameters of $b\approx 5.8\,\mathrm{pc}$ and
$b\approx 3.3\,\mathrm{pc}$ for the $e_0=0.90$ and $e_0=0.99$ cases, respectively.

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
Otherwise the model appears to give a very good match to the simulation results,
and the required slight shift could be due to the galactic potential being slightly tilted in the simulations 
compared to the BH trajectories due to the complex dynamics at this stage.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta \gtrsim 120\degr$ region,
although the model does appear to capture the mean behavior fairly well.
The behavior of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.

Overall we can however see that this simple model correctly captures the behavior of eccentricity seen in the simulations.
On the other hand, if the model is run with low values of $e_\mathrm{s}$,
the binary always reaches extremely high eccentricities with little dependence on the deflection angle,
indicating that the asymmetry of the galactic potential during the SMBH binary formation is the cause for the observed eccentricity behavior.
